Rigidity of Expectations: Additional Evidence from Density Forecasts of the Inflation by Professionals and Households

### 2019-09-12 paper/Draft.tex
Added two citations in literature

%opening
\title{Rigidity of Expectations: Additional Evidence from Density Forecasts of Professionals and Households}
\author{Tao Wang \footnote{Department of Economics, Johns Hopkins University; email: twang80@jhu.edu.}}

@@

	Theories on how agents form expectations in ways deviating from rational expectation(RE) have proliferated over the past decade. This has been also accompanied by an increasingly welcoming recognition of the use of surveys by macroeconomists in testing these theories empirically.  On one hand, various theories built upon different micro-foundations produce somewhat similar macro patterns. For instance, the notion of expectation rigidity micro-founded by different theories mostly generates a sluggish response to new information.  On the other hand, there are important and subtle differences in testable predictions from these theories for both individual forecasts and aggregate moments of the forecasts.

	This paper extends this research effort by bringing an additional moment of forecasts into perspective, i.e. uncertainty. Specifically, in this paper, I first derive the testable predictions from various theories on uncertainty and carefully compare the differences mostly obvious seen from uncertainty. Second, empirically, I test and estimate these implications utilizing the recently available probabilistic forecasts from both professional forecasters survey and household survey. Overall, the paper finds additional evidence from uncertainty consistent with the rigidity models. In the same time, the degree of rigidity appears to be shock-specific and bears the difference between households and professional forecasters.

	The contributions of this paper primarily rely upon the additional insights from the forecast uncertainty, that is only available if surveyees are asked to assign their perceived probabilities to a range of values of the variable of forecast\footnote{For an insightful survey on the importance of measuring subjective expectations using probabilistic surveys, see \citet{manski2004measuring}.}. With density forecasts,  the variance of the individual forecast can be estimated as an indicator of the uncertainty. Across different vintages of the forecast, the revision in uncertainty is a direct measure of information gain or the degree of forecasting efficiency. Besides, the dispersion of uncertainty across agents, and its relationship with other moments widely examined by the literature, such as forecast errors and disagreements provide many insights as to how the agent's expectation actually work.

	This paper is related to four strands of literature. First, the recent decade has seen an emergence of the empirical work directly testing and evaluating the various theories on expectations theories using survey data. For instance, \citet{mankiw2003disagreement}, \citet{carroll2003macroeconomic}, \citet{branch2004theory}. More recent examples include \citet{coibion2018firms} on firms' managers. In addition to testing particular sets of theories, there is also a number of papers that show people's expectations are driven by idiosyncratic demographics, cognitive abilities and macroeconomic histories experienced(\citet{malmendier2015learning}, \citet{das2017socioeconomic} and \citet{d2019iq}, etc.). In terms of the methodology, this paper is closest to \citet{coibion2012can} and \citet{fuhrer2018intrinsic}.  However, all of these studies simply rely upon point forecasts instead of density forecast or surveyed uncertainty. This is one theme on which this paper differs from the existing literature.

@@

	Concerning uncertainty, the first simple implication by FIRE is that all individual shares the same degree of uncertainty. The uncertainty about future $y$ simply comes from uncertainty about unrealized shocks between $t$ and $t+h$. With the same model in mind (Equation \ref{AR_process}) and the same information $y_t$, everyone's uncertainty is equal to the weighted sum of the future volatility before its realization. (Equation \ref{VarREIndPop}). In FIRE, there are neither disagreements about mean, nor disagreements about the uncertainty \footnote{This is the same to \citet{jurado2015measuring}'s terminology.}.

@@

	It is easy to show that the average forecast errors are serially correlated(Equation \ref{FESEPop}). For double-checking, setting $\lambda=1$, SE collapses to FIRE, in which there is no serial correlation between forecast errors and it fully responds to newly realized shocks at time $t$.

@@

	Although the sheer magnitudes of the differences between professionals' moments and those of households are so big that a direct comparison of the two seems redundant, their respective within-agent-type correlation serves a ready checking device of some statistical consistency. Figure \ref{UnceratitnyOtherMoments} plots the population uncertainty against realized inflation,  forecast errors, and disagreements in the first, second and third rows, respectively.

@@

	The middle row of Figure \ref{UnceratitnyOtherMoments} looks into the relationship between size of the forecast error and uncertainty. Although according to our benchmark framework in Section \ref{theory}, there is no mathematical correlation between the size of the forecast errors and uncertainty as the former depends on the realized shocks and the later depends on the volatility of the shock, it is worth checking if in the data a greater ex-ante uncertainty implies bigger ex post forecast errors. The correlation coefficients of the two are -0.19, -0.18 and 0.27 for SPF CPI forecasts, SPF PCE forecasts, and SCE's forecasts, respectively. The two indicators turn out to be negatively correlated for professional forecasts. Only households forecasts exhibit such a positive correlation.

	The last row of Figure \ref{UnceratitnyOtherMoments} examines the relationship between disagreement and uncertainty. Many empirical literature in macroeconomics use disagreements and uncertainty as if they are similar concepts\footnote{For instance, \citet{bachmann2013uncertainty} used ex-ante disagreement and ex-post forecast errors as two measures of uncertainty and find that both uncertainty indicators lead to a reduction in real economic activity.}. Such confusion in practice is partly due to the difficulty of finding appropriate measures of uncertainty in the first place. But as we have presented in \ref{theory}, the two are concepts with distinct statistical definitions. The empirical pattern of the two,  as shown in the plots, also confirms that the two are different objects. The correlation coefficients of the two turn out to be negative, -0.31 and -0.37 for SPF forecasts. In stark contrast, households seem to bear a strong correlation of 0.43 between the two moments.

@@

	Another pattern worth discussing in Figure \ref{IQR_Unceratitny} is that there is a notable rise in the dispersion of professional forecasts in the recent 2-3 years, primarily driven by an increase of upper side of the forecast (i.e. 75 percentile forecast increases from 1\% to 2\%). \footnote{This should be interpreted with caution since the disagreements of SPF forecasts shown in the bottom row of Figure \ref{UncertaintyOtherMoments} actually exhibits a gradual decline. } This is consistent with the observation in the top left two graphs of Figure \ref{Unceratitny_Histogram} that the distribution of inflation forecasts in recent years have become flattened.

@@

	\begin{figure}[ht]
		\centering
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Realized Inflation and Uncertainty}
			\label{InfVar}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_CPICore_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_PCECore_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_CPIAU_varSCEM.png}
		\end{subfigure}
	    \vspace{1em}
		\vfill
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Size of Forecast Errors and Uncertainty}
			\label{FEVar}
			\includegraphics[width=0.3\textwidth]{figures/SPFCPI_abFE_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/SPFPCE_abFE_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/SCE_abFE_varSCEM.png}
		\end{subfigure}
	 \vspace{0.5em}
		\vfill
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Disagreement and Uncertainty}
			\label{DisgVar}
			\includegraphics[width=0.3\textwidth]{figures/CORECPI_disg_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/COREPCE_disg_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Q9_disg_varSCEM.png}\\
		\end{subfigure}
		\\
		\begin{flushleft}
			{\footnotesize Note:  From left to right: SPF's  forecasts of core CPI and core PCE, and SCE's forecast of headline CPI. From top to the bottom, uncertainty (red dash) versus realized inflation (blue dot) with correlation coefficient of 0.14, 0.12 and -0.25, respectively; uncertainty (red dash) versus absolute value of forecast errors (blue dot) with correlation coefficient of -0.19, -0.18, 0.27, respectively; uncertainty (red dash) versus  disagreements (blue dot) with correlation coefficient of -0.31, -0.37 and 0.43,repectively. Only the Pearson tests of correlation between disagreement and uncertainty are significant for all.}
		\end{flushleft}
		\caption{Uncertainty and Other Moments}
		\label{UncertaintyOtherMoments}
	\end{figure}

@@

	\subsection{Replicating \citet{coibion2012can}}

@@

	As to the households forecast, the starkly wider confidence band across shocks and moments seen in Figure \ref{SCE_IR} are not that surprising given the wide dispersion of the forecasts moments of a dramatically larger magnitude seen in Figure \ref{UncertaintyOtherMoments}.

>>>>

%opening
\title{Rigidity of Expectations: Additional Evidence from Density Forecasts of the Inflation by Professionals and Households}
\author{Tao Wang \footnote{Department of Economics, Johns Hopkins University; email: twang80@jhu.edu. }}

@@

	Theories on how agents form expectations in ways deviating from rational expectation(RE) have proliferated over the past decade. This has been also accompanied by a welcoming recognition of the usefulness of surveys by macroeconomists in testing these theories empirically.  On one hand, various theories built upon different micro-foundations produce somewhat similar macro patterns. For instance, the notion of expectation rigidity micro-founded by different theories mostly generates a sluggish response to new information.  On the other hand, there are important and subtle differences in testable predictions from these theories for both individual forecasts and aggregate moments of the forecast.

	This paper extends this line of effort by bringing an additional moment of forecasts into perspective, i.e. the uncertainty. Existing work that studies expectation formation utilizing survey data mostly focus on the individual and cross-sectional patterns of the mean forecast of the variable. But there are additional insights from the forecast uncertainty, which is only available if each surveyee is asked to assign their own perceived probabilities to a range of values of the variable\footnote{For a survey on the importance of measuring subjective expectations using probabilistic surveys, see \citet{manski2004measuring}. See \citet{manski2018survey} for the recent development of the discussion.}. For instance, the dispersion of uncertainty across individual forecasters, as that of the average forecasts, are inconsistent with the benchmark full-information rational expectation as the latter assumes that agents agree on the generating process of the data and have the common knowledge of available information. Dynamically, across different vintages of the forecast, the revision in uncertainty is a measure of information gain or the degree of forecasting efficiency. Besides, the relationship between uncertainty with other moments such as forecast errors and disagreements provides a way to check if the surveyed expectations are consistent with certain theories of expectation formation. Using these insights, the paper finds additional evidence from uncertainty consistent with the rigidity models. In the same time, the degree of rigidity appears to be shock-specific and bears the difference between households and professional forecasters.

	There are three major contributions of this paper. On the theoretical side, I explicitly derive the prediction of various theories of expectation formation about the dynamics of uncertainty. For instance,  different theories that entail rigidities in incorporating new information predict the inefficiency in forecast revisions. This paper shows that this inefficiency can not only be seen in the average forecast, but also in the forecast uncertainty. This imposes additional restrictions on parametric configurations that produce the degree of rigidity seen in the data. Specifically, I find that the configurations of parameters in sticky expectation models are more realistic than that in noisy information theories.

	On the empirical side, the contribution is that we provide additional tests of the null of full-information rational expectation using the revision of uncertainty. I first replicate the results of tests based on forecast errors using microdata of households and professionals following the existing literature. Then, I undertake autoregression-based tests for revisions in mean forecasts and uncertainty, respectively. The new result from this paper is that the revision in uncertainty has a serial correlation that is not consistent with the level of forecast efficiency predicted by rational expectation. This provides an additional result that rejects the null of the rational expectation hypothesis.

	In addition, using externally identified shocks to inflation, i.e. productivity shocks, oil price shocks, and monetary policy shocks, I find evidence that the degree of rigidity is shock-specific. In particular, professionals seem to be responsive to monetary policy shocks, while households are more responsive to oil price shocks. Unlike testing the null-hypothesis, the externally identified shocks allow us to look into the dynamic impulse responses of different moments of expectations to the particular shock. This can be directly compared with the predictions of the theories. I follow the framework by \citet{coibion2012can} while making two variations. First, I do not only examine the dynamics of forecast error, disagreement, but also the forecast uncertainty. Second, I include two monetary policy shocks, the shocks to current federal fund rates and future path of the federal funds rate.

	This paper is related to four strands of literature. First, it is realted to a series of empirical work directly testing and evaluating the various theories on expectations theories using survey data. For instance, \citet{mankiw2003disagreement}, \citet{carroll2003macroeconomic}, \citet{branch2004theory}. More recent examples include \citet{coibion2018firms} on firms' managers. In addition to testing particular sets of theories, there is also a number of papers that show people's expectations are driven by idiosyncratic demographics, cognitive abilities and macroeconomic histories experienced(\citet{malmendier2015learning}, \citet{das2017socioeconomic} and \citet{d2019iq}, etc.). In terms of the methodology, this paper is closest to \citet{coibion2012can} and \citet{fuhrer2018intrinsic}.  However, all of these studies simply rely upon point forecasts instead of density forecast or surveyed uncertainty. This is one theme on which this paper differs from the existing literature.

@@

	Concerning uncertainty, the first simple implication by FIRE is that all individual shares the same degree of uncertainty. The uncertainty about future $y$ simply comes from uncertainty about unrealized shocks between $t$ and $t+h$. With the same model in mind (Equation \ref{AR_process}) and the same information $y_t$, everyone's uncertainty is equal to the weighted sum of the future volatility before its realization (Equation \ref{VarREIndPop}). In FIRE, there are neither disagreements about mean, nor disagreements about the uncertainty \footnote{This is the same to \citet{jurado2015measuring}'s terminology.}.

@@

	It is easy to show that the average forecast errors are serially correlated (Equation \ref{FESEPop}). For double-checking, setting $\lambda=1$, SE collapses to FIRE, in which there is no serial correlation between forecast errors and it fully responds to newly realized shocks at time $t$.

@@

	Although the sheer magnitudes of the differences between professionals' moments and those of households are so big that a direct comparison of the two seems redundant, their respective within-agent-type correlation serves a ready checking device of some statistical consistency. Figure \ref{InfVar}, \ref{FEVar}, and \ref{DisgVar} plot the population uncertainty against realized inflation,  forecast errors, and disagreements in the first, second and third rows, respectively.

@@

	Figure \ref{FEVar} looks into the relationship between size of the forecast error and uncertainty. Although according to our benchmark framework in Section \ref{theory}, there is no mathematical correlation between the size of the forecast errors and uncertainty as the former depends on the realized shocks and the later depends on the volatility of the shock, it is worth checking if in the data a greater ex-ante uncertainty implies bigger ex post forecast errors. The correlation coefficients of the two are -0.19, -0.18 and 0.27 for SPF CPI forecasts, SPF PCE forecasts, and SCE's forecasts, respectively. The two indicators turn out to be negatively correlated for professional forecasts. Only households forecasts exhibit such a positive correlation.

	Figure \ref{DisgVar} examines the relationship between disagreement and uncertainty. Many empirical literature in macroeconomics use disagreements and uncertainty as if they are similar concepts\footnote{For instance, \citet{bachmann2013uncertainty} used ex-ante disagreement and ex-post forecast errors as two measures of uncertainty and find that both uncertainty indicators lead to a reduction in real economic activity.}. Such confusion in practice is partly due to the difficulty of finding appropriate measures of uncertainty in the first place. But as we have presented in \ref{theory}, the two are concepts with distinct statistical definitions.\footnote {This point was made very clearly by \citet{zarnowitz1987consensus}.  \citet{manski2018survey} also points out many empirical work has confused the dispersion with uncertainty.} The empirical pattern of the two,  as shown in the plots, also confirms that the two are different objects. The correlation coefficients of the two turn out to be negative, -0.31 and -0.37 for SPF forecasts. In stark contrast, households seem to bear a strong correlation of 0.43 between the two moments.

@@

	Another pattern worth discussing in Figure \ref{IQR_Unceratitny} is that there is a notable rise in the dispersion of professional forecasts in the recent 2-3 years, primarily driven by an increase of upper side of the forecast (i.e. 75 percentile forecast increases from 1\% to 2\%). \footnote{This should be interpreted with caution since the disagreements of SPF forecasts shown in Figure \ref{DisgVar} actually exhibits a gradual decline. } This is consistent with the observation in the top left two graphs of Figure \ref{Unceratitny_Histogram} that the distribution of inflation forecasts in recent years have become flattened.

@@

	\begin{figure}[ht]
		\centering
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Realized Inflation and Uncertainty}
			\label{InfVar}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_CPICore_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_PCECore_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Inf1yf_CPIAU_varSCEM.png}
		\end{subfigure}
	    \vspace{1em}
		\vfill
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Size of Forecast Errors and Uncertainty}
			\label{FEVar}
			\includegraphics[width=0.3\textwidth]{figures/SPFCPI_abFE_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/SPFPCE_abFE_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/SCE_abFE_varSCEM.png}
		\end{subfigure}
	 \vspace{0.5em}
		\vfill
		\begin{subfigure}[b]{\textwidth}
			\centering
			\caption{Disagreement and Uncertainty}
			\label{DisgVar}
			\includegraphics[width=0.3\textwidth]{figures/CORECPI_disg_varSPFCPIQ.png}
			\includegraphics[width=0.3\textwidth]{figures/COREPCE_disg_varSPFPCEQ.png}
			\includegraphics[width=0.3\textwidth]{figures/Q9_disg_varSCEM.png}\\
		\end{subfigure}
		\\
		\begin{flushleft}
			{\footnotesize Note:  From left to right: SPF's  forecasts of core CPI and core PCE, and SCE's forecast of headline CPI. From top to the bottom, uncertainty (red dash) versus realized inflation (blue dot) with correlation coefficient of 0.14, 0.12 and -0.25, respectively; uncertainty (red dash) versus absolute value of forecast errors (blue dot) with correlation coefficient of -0.19, -0.18, 0.27, respectively; uncertainty (red dash) versus  disagreements (blue dot) with correlation coefficient of -0.31, -0.37 and 0.43,repectively. Only the Pearson tests of correlation between disagreement and uncertainty are significant for all.}
		\end{flushleft}
		\caption{Uncertainty and Other Moments}
	\end{figure}

@@

	\subsection{Forecast errors and disagreement}

@@

	As to the households forecast, the starkly wider confidence band across shocks and moments seen in Figure \ref{SCE_IR} are not that surprising given the wide dispersion of the forecasts moments of a dramatically larger magnitude seen in Figure \ref{IQR_Unceratitny}.
